Predictive Quantization for MIMO-OFDM SVD Precoders using Reservoir Computing Framework

### 2019-04-30 paper/main.tex
Changes in result section

\section{Introduction}
\label{intro}
% Broad summary of your work, come to your problem definition.
In MIMO wireless systems, precoding matrices are used for transformation of an $N_S$-dimensional ($N_S \leq \text{min}(N_T,N_R))$ information vector onto an $N_T$-dimensional transmit vector that corresponds to the signal emanating from the $N_T$ antennas of the transmitter.
The precoding matrix used to achieve the channel capacity using optimal power allocation is obtained using the SVD of the MIMO channel matrix \cite{love2008overview}.
The MIMO channel matrix is estimated at the receiver, and the transmitter has no a priori knowledge of the same.
Therefore, to obtain the said benefits of SVD precoders, the receiver needs to quantize and feed back the precoders it obtains from the SVD of the MIMO channel matrix.
To respect the bit budget imposed by limited feedback CSI schemes, the quantization needs to be performed using very few bits.
Given the high dimensionality of the precoders (which form a $N_T \times N_R$ complex valued matrix, viz. $2N_TN_R$ real numbers), effective quantization with just a few bits poses a significant challenge.

% Elaborate on each point (what people have done before) lecunnas
Past work has used two inherent properties of the precoders to enable effective quantization of these precoders, with just a few bits.
First, the precoders are not actually $2N_TN_R$ dimensional entities, and these have an underlying manifold structure that allows one to work with lower dimensions, if one performs operations over the manifold directly.
Second, these precoders are correlated in both time and frequency.
Capturing the available correlations permit the use of predictive quantization algorithms, which enable improvement of the quantization error with time.
This improvement is obtained because exploiting the available correlations from past fed back precoders allow one to just quantize the extra information required on top of the past available temporal-frequency information, instead of the new precoder information in the entirety.
Exploiting frequency correlations permits the use of interpolation algorithms, which in turn reduce the feedback overhead in a single OFDM frame, by feeding back CSI information only for certain subcarriers, and interpolating over the others.
Thus, combining these two inherent properties, viz. the underlying manifold structure and temporal-frequency correlations, requires a generalization of such linear algorithms to the manifold under consideration, by exploiting the differential geometry of the manifold. Relevant past work in this direction has been presented in \cite{Gupt1905:Predictive,6891198,6545375,5671092,Li2016,khaled2005quantized,krishnamachari2013geometry}.

% What is the way ahead
Manifold based approaches for predictive quantization have provided significant performance benefits, given that they work with the effective lower dimensions while framing the manifold operations to predict and quantize.
However, these approaches are difficult to implement and require some operations on the manifold structure.
In this paper, we propose a reservoir computing based predictive quantizer to exploit the temporal correlations effectively.
The proposed scheme offers a simple solution to the predictive quantization problem, which brings about both ease of implementation, and improved performance as well, when compared to \cite{6891198}, which proposed a manifold based method of predictive quantization by exploiting temporal correlations.

@@

\noindent Reservoir computing has also been applied to solve problems in the domain of wireless communications \cite{jaeger2004harnessing,mosleh2017brain,shafin2018realizing}.
\cite{jaeger2004harnessing} used reservoir computing framework to harness channel non-linearities for smarter channel equalization.
\cite{mosleh2017brain,shafin2018realizing} utilized the framework for OFDM symbol detection while accounting for channel and power amplifier non-linearities.
In our work, we utilize the reservoir computing framework to capture the underlying nonlinear relations between previously observed precoders, in order to predict the precoder at next time instant. This becomes analogous to time series prediction algorithms, for which the reservoir computing framework has been known to work effectively \cite{mosleh2017brain}.

\section{System Model}
\label{section2}
We consider a point-to-point $N_T\times N_R$ MIMO-OFDM wireless system.
The available bandwidth is divided into $N$ subcarriers, so that, individually for each subcarrier, the channel can be assumed to be flat fading.
In this discussion, we assume that $N_T > N_R$ and $N_s = N_R$. Consistent with notation in \cite{6891198,Gupt1905:Predictive}, the data stream received on the $i$-th subcarrier is denoted by:
\begin{equation}
\by_{i,t}=\bH^{H}_{i,t}\tilde{\bU}_{i,t}\bx_{i,t}+ \bw_{i,t}
\end{equation}
$\tilde{\bU}_{i,t} \in \mathbb{C}^{N_T \times N_R}$ represents the precoding matrix, which is a function of the quantized CSI fed back by the receiver, $\by_{i,t} \in \mathbb{C}^{N_R \times 1}$ is the received data stream, $\bx_{i,t} \in \mathbb{C}^{N_T \times 1}$ denotes the transmitted signal, $\bH_{i,t} \in \mathbb{C}^{N_T \times N_R}$ denotes the MIMO channel matrix and $\bw_{i,t}$ denotes the i.i.d. complex Gaussian noise with $\bw_{i,t} \sim \mathcal{N}_{\mathbb{C}}(0,N_0\bI_{N_R})$, $N_0$ being the noise variance.

@@

Independent quantization algorithms analogous to the Lloyd codebook algorithm have been studied for quantizing $\text{St}(N_T,N_R)$ \cite{6678348}.
However, this approach treats precoders that are close to each other in time/frequency as independent, since it does not exploit any temporal/spectral correlations for quantization.
Predictive quantization algorithms \cite{Gupt1905:Predictive,6891198} can be utilized to reduce the quantization error by utilizing the available correlations.
We now discuss the basic mathematical model that is central to predictive quantization algorithms for precoding matrices.
Suppose that we predict the current precoding matrix based on the past $p$ observed precoding matrices, and hence $\tilde{\bU}_{i,t-p},\tilde{\bU}_{i,t-p+1},\ldots,\tilde{\bU}_{i,t-2},\tilde{\bU}_{i,t-1}$ have been fed back and available at the transmitter.
By using a prediction algorithm and by exploiting time correlations among precoders at time instances $t-p,t-p+1,\ldots,t-2,t-1$, the transmitter can obtain a prediction (coarse estimate) of $\bU_{i,t}$, as $\bP_{i,t}=f_P(\tilde{\bU}_{i,t-p},\tilde{\bU}_{i,t-p+1},\ldots,\tilde{\bU}_{i,t-2},\tilde{\bU}_{i,t-1})$, where $f_P: \text{St}(N_T,N_R)\times\ldots\{p-\text{times}\}\ldots\times\text{St}(N_T,N_R) \to \text{St}(N_T,N_R)$ is a prediction function.
Now, since the transmitter already possesses a coarse estimate $\bP_{i,t}$ of $\bU_{i,t}$, the receiver just quantizes the local space of $\text{St}(N_T,N_R)$ nearby $\bP_{i,t}$, instead of the complete $\text{St}(N_T,N_R)$. The receiver feeds back this quantized information to the transmitter, enabling estimation of a refined estimate of $\bU_{i,t}$, given by $\tilde{\bU}_{i,t}$.
That is, the transmitter utilizes a quantization function $q_P:\text{St}(N_T,N_R) \times \mathbb{N} \to \text{St}(N_T,N_R)$, to obtain $\tilde{\bU}_{i,t} \gets q_P(\bP_{i,t},\text{fb}_{i,t})$, where $\text{fb}_{i,t} \in \{1,2\ldots2^{\mathcal{B}}\}$ is the feedback from the receiver for precoder at $i$-th subcarrier and $t$-th time instant, considering bit budget of $\mathcal{B}$ bits.
In \cite{Gupt1905:Predictive,6891198,6545375}, $\text{fb}_{i,t}$ denotes the codeword index for the codebook obtained for quantizing the local tangent space at $\bP_{i,t}$.

When using the predictive quantization scheme, the quantization is performed on a smaller subspace when compared to the entire manifold in case of independent quantization.
Therefore, the quantization error is substantially lower as well.
However, independent quantization algorithms are important for smart initialization (instead of a cold start) of the predictive quantization strategies.
Observe that predictive quantization algorithms for SVD precoders are typically the higher dimensional analogues of well known approaches to reduce quantization error using linear prediction codes (LPCs) and Delta PCM.
In both LPCs and Delta PCM, quantization error can be reduced substantially using a linear predictor and quantizing the scalar difference between the predicted and observed value (innovation).
Analogous to this, for $\text{St}(N_T,N_R)$, quantization error is reduced by quantizing a subspace around the predicted value that is much smaller than the entire manifold.

@@

We simulate the (uncoded QPSK) BER performance for precoders obtained after interpolation ($\tilde{\bU}_{i\neq\{9k\},t}$) and quantization ($\tilde{\bU}_{\{9k\},t}$), for all the 63 subcarriers. Observe from Fig. \ref{BER} that reservoir computing framework is able to achieve substantial improvements in $\text{E}_\text{b}\text{N}_0$ levels at $\text{BER}\leq 10^{-3}$. Particularly, observe, for $\text{BER}=10^{-4}$ we observe around $5$ dB improvement. Also, notice the substantial improvement which both the predictive quantization schemes offer from the independent quantization scheme with Lloyd codebook for precoders \cite{6678348}, using the same number of bits.

>>>>

\section{Introduction}
\label{intro}
% Broad summary of your work, come to your problem definition.
In MIMO wireless systems, precoding matrices are used for transformation of an $N_S$-dimensional ($N_S \leq \text{min}(N_T,N_R))$ information vector onto an $N_T$-dimensional transmit vector that corresponds to the signal emanating out of the $N_T$ antennas of the transmitter.
The precoding matrix used to achieve the channel capacity using optimal power allocation is obtained using the SVD of the MIMO channel matrix \cite{love2008overview}.
The MIMO channel matrix is estimated at the receiver, and the transmitter has no a priori knowledge of the same.
Therefore, to obtain the said benefits of SVD precoders, the receiver needs to quantize and feed back the precoders it obtains from the SVD of the MIMO channel matrix.
To respect the bit budget imposed by limited feedback CSI schemes, the quantization needs to be performed using very few bits.
Given the high dimensionality of the precoders (which form a $N_T \times N_R$ complex valued matrix, viz. $2N_TN_R$ real numbers), effective quantization with just a few bits poses a significant challenge.

% Elaborate on each point (what people have done before) lecunnas
Past work has used two inherent properties of the precoders to enable effective quantization of these precoders, with just a few bits.
First, the precoders are not actually $2N_TN_R$ dimensional entities, and these have an underlying manifold structure that allows one to work with lower dimensions, if one performs operations over the manifold directly.
Second, these precoders are correlated in both time and frequency.
Capturing the available correlations allow for predictive quantization algorithms, which enable improvement of the quantization error with time.
This improvement is obtained because exploiting the available correlations from past fed back precoders allow one to just quantize the extra information required on top of the past available temporal-frequency information, instead of the new precoder information in the entirety.
Exploiting frequency correlations permits the use of interpolation algorithms, which in turn reduces the feedback overhead in a single OFDM frame, by feeding back CSI information only for certain subcarriers, and interpolating over the others.
Thus, combining these two inherent properties, viz. the underlying manifold structure and temporal-frequency correlations, requires a generalization of such linear algorithms to the manifold under consideration, by exploiting the differential geometry of the manifold. Relevant past work in this direction has been presented in \cite{Gupt1905:Predictive,6891198,6545375,5671092,Li2016,khaled2005quantized,krishnamachari2013geometry}.

% What is the way ahead
Manifold based approaches for predictive quantization have provided significant performance benefits, given that they work with the effective lower dimensions while framing the manifold operations to predict and quantize.
However, these approaches are difficult to implement and require some familiarity with manifold geometry for a complete grasp.
In this paper, we propose a reservoir computing based predictive quantizer to exploit the temporal correlations effectively.
The proposed scheme offers a cleaner solution to the predictive quantization problem, which brings about both ease of implementability, and improved performance as well, when compared to \cite{6891198}, which proposed a manifold based method of predictive quantization by exploiting temporal correlations.

@@

\noindent Reservoir computing has also been applied to solve problems in the domain of wireless communications \cite{jaeger2004harnessing,mosleh2017brain,shafin2018realizing}.
\cite{jaeger2004harnessing} used reservoir computing framework to harness channel non-linearities for smarter channel equalization.
\cite{mosleh2017brain,shafin2018realizing} utilized the framework for OFDM symbol detection while accounting for channel and power amplifier non-linearities.
In our work, we utilize the reservoir computing framework to discover the underlying nonlinear relations between previously observed precoders, in order to predict the precoder at next time instant. This becomes analogous to time series prediction algorithms, for which the reservoir computing framework has been known to work effectively \cite{mosleh2017brain}.

\section{System Model}
\label{section2}
\noindent We consider a point-to-point $N_T\times N_R$ MIMO-OFDM wireless system.
The available bandwidth is divided into $N$ subcarriers, so that, individually for each subcarrier, the channel can be assumed to be flat fading. In this discussion, we assume that $N_T > N_R$ and $N_s = N_R$. Consistent with notation in \cite{6891198,Gupt1905:Predictive}, the data stream received on the $i$-th subcarrier is denoted by:
\begin{equation}
\by_{i,t}=\bH^{H}_{i,t}\tilde{\bU}_{i,t}\bx_{i,t}+ \bw_{i,t}
\end{equation}
$\tilde{\bU}_{i,t} \in \mathbb{C}^{N_T \times N_R}$ represents the precoding matrix, which is a function of the quantized CSI fed back by the receiver, $\by_{i,t} \in \mathbb{C}^{N_R \times 1}$ is the received data stream, $\bx_{i,t} \in \mathbb{C}^{N_T \times 1}$ denotes the transmitted signal, $\bH_{i,t} \in \mathbb{C}^{N_T \times N_R}$ denotes the MIMO channel matrix and $\bw_{i,t}$ denotes the i.i.d. complex Gaussian noise with $\bw_{i,t} \sim \mathcal{N}_{\mathbb{C}}(0,N_0\bI_{N_R})$, $N_0$ being the noise variance.

@@

Independent quantization algorithms analogous to the Lloyd codebook algorithm have been studied for quantizing $\text{St}(N_T,N_R)$ \cite{6678348}.
However, this approach treats precoders that are close to each other in time/frequency as independent, since it does not exploit any temporal/spectral correlations for quantization.
Predictive quantization algorithms \cite{Gupt1905:Predictive,6891198} can be utilized to reduce the quantization error by utilizing the available correlations. We now discuss the basic mathematical model that is central to predictive quantization algorithms for precoding matrices.
Suppose that we predict the current precoding matrix based on the past $p$ observed precoding matrices, and hence $\tilde{\bU}_{i,t-p},\tilde{\bU}_{i,t-p+1},\ldots,\tilde{\bU}_{i,t-2},\tilde{\bU}_{i,t-1}$ have been fed back and available at the transmitter.
By using a prediction algorithm and by exploiting time correlations among precoders at $t-p,t-p+1,\ldots,t-2,t-1$ time instances, the transmitter can obtain a prediction (coarse estimate) of $\bU_{i,t}$, as $\bP_{i,t}=f_P(\tilde{\bU}_{i,t-p},\tilde{\bU}_{i,t-p+1},\ldots,\tilde{\bU}_{i,t-2},\tilde{\bU}_{i,t-1})$, where $f_P: \text{St}(N_T,N_R)\times\ldots\{p-\text{times}\}\ldots\times\text{St}(N_T,N_R) \to \text{St}(N_T,N_R)$ is a prediction function.
Now, since the transmitter already possesses a coarse estimate $\bP_{i,t}$ of $\bU_{i,t}$, the receiver just quantizes the local space of $\text{St}(N_T,N_R)$ nearby $\bP_{i,t}$, instead of the complete $\text{St}(N_T,N_R)$. The receiver feeds back this quantized information to the transmitter, enabling estimation of a refined estimate of $\bU_{i,t}$, given by $\tilde{\bU}_{i,t}$.
That is, the transmitter utilizes a quantization function $q_P:\text{St}(N_T,N_R) \times \mathbb{N} \to \text{St}(N_T,N_R)$, to obtain $\tilde{\bU}_{i,t} \gets q_P(\bP_{i,t},\text{fb}_{i,t})$, where $\text{fb}_{i,t} \in \{1,2\ldots2^{\mathcal{B}}\}$ is the feedback from the receiver for precoder at $i$-th subcarrier and $t$-th time instant, considering bit budget of $\mathcal{B}$ bits.
In \cite{Gupt1905:Predictive,6891198,6545375}, $\text{fb}_{i,t}$ denotes the codeword index for the codebook obtained for quantizing the local tangent space at $\bP_{i,t}$.

When using predictive quantization scheme, the quantization is performed on a smaller subspace when compared to the entire manifold in case of independent quantization.
Therefore, the quantization error is substantially lower as well.
However, independent quantization algorithms are important for smart initialization (instead of a cold start) of the predictive quantization strategies.
Observe that predictive quantization algorithms for SVD precoders are typically the higher dimensional analogues of the well known approaches to reduce quantization error using linear prediction codes (LPCs) and Delta PCM.
In both LPCs and Delta PCM, quantization error can be reduced substantially using a linear predictor and quantizing the scalar difference between the predicted and observed value (innovation).
Analogous to this, for $\text{St}(N_T,N_R)$, quantization error is reduced by quantizing a subspace around the predicted value that is much smaller than the entire manifold.

@@

We simulate the (uncoded QPSK) BER performance for precoders obtained after interpolation ($\tilde{\bU}_{i\neq\{9k\},t}$) and quantization ($\tilde{\bU}_{\{9k\},t}$), for all the 63 subcarriers. Observe from Fig. \ref{BER} that reservoir computing framework is able to achieve substantial improvements in $\text{E}_\text{b}\text{N}_0$ levels at $\text{BER}\leq 10^{-4}$. Particularly, observe, for $\text{BER}=10^{-5}$ we observe around $5$ dB improvement.
